First and foremost I would like to thank my advisor, Andrew MacFadyen, for his support and endless encouragement.  Andrew is a font of ideas and enthusiasm, if I can channel just a fraction of his energy I know my career will be a success.

My thesis committee, Alexander Grosberg, Andrei Gruzinov, David Hogg, and Maryam Modjaz, have all been hugely influential on my scientific career.  Through classes, chats, and a great deal of coffee their advice and attitude have steered me throughout my time at NYU.  

Former group members Paul Duffell, Brian Farris, Hendrik van Eerten, and Jonathan Zrake taught me everything I know.  They lent me their ears, and more importantly their code. 

Great thanks to Patrick Cooper: colleague and friend.  My officemates Chang, Nitya, Kilian, MJ, Alex, Oliver, and Rob have put up with much and talked me through more, I only hope I was as supportive for them as they have been for me.  Current group members Yike Tang, Xiaoyi Xie, and Yiyang Wu have put up with me barging into their offices for years.  Cato Sandford, colleague, friend, roommate, and purveyor of the finest craic. In September of 2011, nineteen people entered the NYU Physics grad program and found themselves kettled in one giant office.  What could have been bloody ended with us all becoming the best of friends.  To those of my cohort I haven't mentioned yet, Ali, Chen, Christine, Doycho, Hossein, Jie, Mark, Ruben, Yin, Yu-Ming, and Yuqian, thank you for the adventures and your friendship.

My family has been a pillar of support and encouragement since the first day I watched Back to the Future and decided I wanted to be Doc Brown when I grew up.  Mom, Dad, Eric, Mark, and Kathy, I would not be here without you.  

Thank you Naomi Lewandowski, my partner in this mad adventure.